Counting evolutions: a simple foundation for thermodynamics

### 2021-02-10 Articles/2019-Entropy/Source/thermodynamics.tex
Another pass at the thermodynamics article

We now want to return to the link between evolution count, state entropy and state spaces. In particular, we want to show how this link exists not only in thermodynamics, but also in classical mechanics, quantum mechanics and statistical mechanics. This is important not only to show how process entropy relates to the two main ways to define entropy in statistical mechanics, but to give a sense that all geometrical and probabilistic structure assigned to state spaces are ultimately capturing evolution counts in one way or another. This is important for our bigger project, Assumptions of Physics, as we want to create a common framework for all physical theories.

We previously defined mechanical systems as those for which all states have the same entropy. While one does not usually discuss the entropy of single classical or quantum states, it is easy to see that that must be the case. In quantum mechanics, for example, the von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to count states, which means all states must contribute the same entropy. Additionally, in both theories, given two states, we can always (at least mathematically) find a deterministic and reversible process (Hamiltonian evolution and unitary evolution are entropy preserving) that starts with the first and ends in the second. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, is already embedded in the established theories.

The geometrical structure of classical mechanics, the symplectic structure of phase space, can be understood to be exactly the structure we need to count evolutions when the evolution is deterministic and reversible and the degrees of freedom are independent. In that case state entropy and process entropy coincide, but not in general. In dissipative processes, like a damped harmonic oscillator, this relationship breaks down. As the system loses energy, the evolutions gradually concentrate around the attractors. In terms of count of states, areas shrink over time, so counting states gives us a decreasing number, which is at odds with the idea of entropy increasing when progressing towards an equilibrium. What happens is that as the system evolves, each region will correspond to the same number of evolutions, which will now be spread over fewer states: the process entropy (i.e. the logarithm of the evolutions per state) increases as evolutions become more concentrated.

The mathematical structure of quantum mechanics also can be understood in terms of evolution counts. The inner product of quantum mechanics gives us the conditional probabilities of a ``black box'' process (i.e. the projection/measurements) that, given an initial state, gives us a final one. As we saw, conditional probabilities are ratios of evolution counts: what fractions of the evolutions that start in one state end in another. In other words, the inner product tells us how the evolutions coming in and coming out of these processes are to be ``joined'' with the preceding and following ones.  Moreover, note that these processes are projections, meaning they give the same result if applied more than once. Eigenstates of projections, then, are equilibria of the process the projection represents. Or, in other words, equilibria are symmetries of the process.

In statistical mechanics, one uses the above structures to define the entropy of an ensemble. There are currently two ways to do this. The first, commonly referred to as the fundamental postulate of statistical mechanics, is to take the logarithm of the state count of the region of phase space identified by the constraints (e.g. the energy of the system is within a particular bound). We can relate this to the process entropy in the following way. Suppose that the constraints are such that each microstate either is compatible or incompatible with the macrostate. Then the count of evolutions of the macrostate is the sum of all the evolutions of its microstates. Now suppose that, when the constraints are defined, the system is at equilibrium and independent. Then the process entropy is equal to the state entropy. Since the system is mechanical, the state entropy is the same for all states. This means that, if the conditions above are met, the process entropy is, up to a constant, the logarithm of the state count.\footnote{If the system starts independent and at equilibrium, for example, the entropy of a final equilibrium would correspond to the logarithm of the size of the basin of attraction: each state corresponds to a single evolution that ends in the attractor.}

The second way to define entropy is through the Gibbs and von Neumann entropies, which essentially are the application of the Shannon entropy in the classical and quantum case. This can be related to the process entropy in the following way. Suppose we have a macroscopic equilibrium which can be understood as a dynamical equilibrium of the microscopic description. That is, the macroscopic equilibrium is a set of evolutions whose microscopic description can be characterized by a stable probability distribution $\rho(x)$. An evolution $\lambda(t)$ for the microstate, then, is an dense sequence of infinitely many microstates whose recurrence matches $\rho(x)$. Since the state entropy is the maximal process entropy, we are counting all possible evolutions which means all possible permutations. The Shannon entropy can be understood as the count of all possible permutations of an infinite sequence, of all possible evolutions.

The above insights show that the framework we are setting up is not just a nicer retelling of thermodynamics, but it is something more fundamental that underlies all physical theories. Most, if not all, experiments are described as a sequence of different steps (e.g. one or more preparation, manipulation and measurement) stitched together. Making sure that the evolution count matches at every junction is a very simple and necessary bookkeeping requirement, and therefore it makes sense it plays such a major role in defining state spaces.

Therefore it makes sense that countMaking sure that the evolution count  Stitching together those phases into a single process

The idea of using states as junctures when processes change is at the core of scientific experimentation since most experiments are

 maostate preparation)can help us better understand the fundamental connection between all these theories. We stress that we are not arguing that all other notions of entropy should be abandoned in favor of process entropy: quite the contrary. Specific tools are often more useful than general tools precisely because they take advantage of the specific nature of the problem. The point is that if we are more aware of the interplay between the general definition and the more specific one we will not use the latter inappropriately.

@@

In this paper we gave a conceptual overview for a new approach to recover thermodynamics based on the simple idea of counting evolutions, the possible ways a system can evolve under a process. We saw that the logarithm of evolution count under a deterministic process has all the properties we associate with entropy. We have seen that the evolution count is maximized at equilibrium and for independent system and only in those cases it depends only on the state. A well defined evolution count is necessary if we want to be able to compose processes, since the incoming evolutions needs to match the outgoing evolutions for all states at the junction. Thermodynamics studies exactly the case of composite processes of composite systems under some simplifying conditions. This gives a simple way to rederive thermodynamics in a straight-forward and conceptually insightful way.

Moreover, we have seen that the state spaces of classical, quantum and statistical mechanics include geometrical and probabilistic structures that can all be understood in terms of evolution count. Our hope and long term goal is to develop a formal theory based on these ideas that identifies a minimum set of requirements for the definition of a state space. This theory can be further refined with additional assumptions in different ways to recover the different physical theories. We believe this would help us better understand the right questions to ask in within the foundations of physics.

>>>>

We now briefly explore the link between evolution count, state entropy and state spaces in other physical theories, namely classical mechanics, quantum mechanics and statistical mechanics. This is important not only to show how process entropy relates to the two main ways to define entropy in statistical mechanics, but to give a sense that all geometrical and probabilistic structures assigned to state spaces are ultimately capturing evolution counts in one way or another. This is important for our bigger project, Assumptions of Physics, as we want to create a common framework for all physical theories.

In statistical mechanics there are currently two widely accepted ways to define the entropy of an ensemble. The first, commonly referred to as the fundamental postulate of statistical mechanics, is to take the logarithm of the state count of the region of phase space identified by the constraints (e.g. the energy of the system where some quantities are within a particular bound). One way we can relate this to the process entropy is the following. Suppose that the dynamics is Hamiltonian and suppose the constraints are preserved by the dynamics. Since each microstate provides a full description of the system, all microstates correspond to the same number of evolutions. Therefore the process entropy of the macrostate is, up to a constant, the logarithm of the count of states that satisfy the constraint.\footnote{This is not the only way to relate the two, but it best corresponds to the typical setup one finds in the literature.}

The second way to define entropy is through the Gibbs and von Neumann entropies, which essentially are the application of the Shannon entropy in the classical and quantum case. One way we can relate this to the process entropy is the following. Suppose we have a macroscopic equilibrium which can be understood as a dynamical equilibrium of the microscopic description. That is, the macroscopic equilibrium is a set of evolutions whose microscopic description can be characterized by a stable probability distribution $\rho(x)$. An evolution $\lambda(t)$ for the microstate, then, is an dense sequence of infinitely many microstates whose recurrence matches $\rho(x)$. Since the state entropy is the maximal process entropy, we are counting all possible evolutions which means all possible permutations. The Shannon entropy can be understood as the count of all possible permutations of an infinite sequence, of all possible evolutions.

These links are very straight forward and are fundamental in the sense that we are not applying different definitions that happen to agree: we are trivially applying the general concept, counting evolutions, in the different cases and finding the correct result. There is no mystery as to what is going on in the above two paragraphs, it's just a computation.

Note, however, that both definition relies on a somewhat implicit premise: all classical and quantum states individually have the same state entropy, they are, as we defined them, mechanical systems. In quantum mechanics, for example, the von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to count states, which means all states must contribute the same entropy. Additionally, in both theories, given two states, we can always (at least mathematically) find a deterministic and reversible process (Hamiltonian evolution and unitary evolution are entropy preserving) that starts with the first and ends in the second. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, is already embedded in the established theories. In fact, all geometrical features of those spaces can be understood in terms of counting evolutions.

The geometrical structure of classical mechanics (i.e. the symplectic structure of phase space) can be understood to be exactly the structure we need to count the evolutions of a deterministic and reversible process (i.e. a Hamiltonian process). In that case state entropy and process entropy coincide, but not in general. For example, in dissipative processes (e.g. damped harmonic oscillator), this relationship breaks down. As the system loses energy, the evolutions gradually concentrate around attractors. In terms of count of states, areas shrink over time, so counting states gives us a decreasing number, which is at odds with the idea of entropy increasing when progressing towards an equilibrium. On the other hand, each region as it evolves will correspond to the same number of evolutions, which will now be spread over fewer states: the process entropy (i.e. the logarithm of the evolutions per state) increases as evolutions become more concentrated. In the same way that thermodynamic entropy is the process entropy under isoentropic evolution (i.e. at equilibrium), the state entropy in classical mechanics is the process entropy under isoentropic evolution (i.e. deterministic and reversible evolution). Therefore, in the same way thermodynamic entropy does not apply outside of equilibrium, one has to be careful to apply the symplectic structure outside of Hamiltonian evolution.

The mathematical structure of quantum mechanics also can be understood in terms of evolution counts. The inner product of quantum mechanics gives us the conditional probabilities of a ``black box'' process (i.e. the projection/measurements) that, given an initial state, gives us a final one. As we saw, conditional probabilities are ratios of evolution counts: what fractions of the evolutions that start in one state end in another. In other words, the inner product tells us how the evolutions coming in and coming out of these processes are to be ``joined'' with the preceding and following ones.  Unitary evolution is exactly when all evolution of one state go to one and only one state.  Moreover, note that these ``black box'' processes are projections, meaning they give the same result if applied more than once. Eigenstates of projections, then, are equilibria of the process the projection represents. Or, in other words, equilibria are symmetries of the process.

The above insights show that the concepts we are using do not just provide a nicer retelling of thermodynamics, but they relate to something more fundamental that underlies all physical theories. We believe that this is not a coincidence, but stems from an obvious requirement that we must put on physical theories. Most, if not all, experiments are described as a sequence of different steps (e.g. one or more preparation, manipulation and measurement) stitched together. Therefore a theory must be able to tell us how to ``stitch together'' the evolutions at every junction, how to keep the evolution count correctly. This bookkeeping is a requirement and therefore it makes sense it plays such a major role in defining state spaces.

@@

In this paper we gave a conceptual overview for a new approach to recover thermodynamics based on the simple idea of counting evolutions, the possible ways a system can evolve under a process. We saw that the logarithm of evolution count under a deterministic process has all the properties we associate with entropy: it cannot decrease over time, it is constant in reversible processes and it is maximized at equilibrium. The state entropy represents the maximized process entropy at equilibrium and for the composite of independent systems. Thermodynamic processes and equilibria describe the case where the equilibrium is achieved at all levels: the whole and any of the parts. This gives a simple way to rederive thermodynamics in a straight-forward and conceptually insightful way.

We have also seen that two main ways to define entropy in statistical mechanics can be understood as the process entropy for the ensembles, and that the geometrical properties of classical and quantum mechanics can be understood in terms of evolutions counts. Keeping track of evolution counts is fundamental in a physical theory for process composition, as we need to make sure that the evolutions are ``stitched together'' correctly when processes are joined. Our hope and long term goal is to develop a formal theory based on these ideas that identifies a minimum set of requirements for the definition of a state space. This basic definition can then be further refined with different additional assumptions, each set recovering the different physical theories. We believe this would help us better understand the right questions to ask in within the foundations of physics.
